---
title: "Introducing Patch Preference to Levins Metapopulation Model (1969)"
author: Martina Reina Canitrot (2215336)
output: html_document
date: "22-04-2024"
---

```{r setup, include=FALSE}
chooseCRANmirror(ind = 1)
install.packages("knitr")
knitr::opts_chunk$set(echo = TRUE)
```

<span style="font-size: large;">Introducing Patch Preference to Levins Metapopulation Model (1969).</span>

<span style="font-size: medium;">This document features a version of the Levins metapopulation model, originally aiming to simulate the rate of change in metapopulation size, measured by the fraction of occupied patches at the time.</span>

The Metapopulation Patch Preference model assumes:  

1.	Patches are connected allowing dispersal of individuals between them, either physically (i.e. stepping stones) or functionally (i.e. migration, seed dispersal).
  a.	This is influenced by preference parameter (v) influencing the           dispersal patter of individuals.

2.	Density-dependent effects on population growth within each patch. Where growth rate of population changes as a function of its population density

3.	Parameters c, e and T remain constant over time and patches when v is changing.

Model Equation:

$\frac{dN}{dt} = cN\left(1-\frac{N}{T}+\frac{T-N}{T} \cdot \frac{1}{v+1}\right)-eN$

The model code:
```{r}
#dN <- c * N * (1 - N/T - (T-N)/T * (1/(v+1))) - e * N
```


Download the necessary packages 
```{r}
install.packages("ggplot2")
library(ggplot2)
install.packages("deSolve")
library(deSolve)
```

The diagrams below are used as visual illustrations of theoretical patch dynamics showing the difference between patch dynamics between the original Levins model (1969) and our reformed patch preference model shown above, where dispersers may favour one patch over the other due to more favourable conditions.


### Random Dispersal
```{r}
#DIAGRAM RANDOM DISPERSAL
patches <- data.frame(
  patch = c("Patch 1", "Patch 2", "Patch 3"),
  x = c(1, 2, 3),
  y = c(1, 3, 2)
)

movement <- data.frame(
  from = c("Patch 1", "Patch 1", "Patch 2", "Patch 2", "Patch 3"),
  to = c("Patch 2", "Patch 3", "Patch 1", "Patch 3", "Patch 1")
)

movement_coords <- merge(movement, patches, by.x = "from", by.y = "patch")
movement_coords <- merge(movement_coords, patches, by.x = "to", by.y = "patch", suffixes = c("_from", "_to"))

ggplot() +
  geom_point(data = patches, aes(x = x, y = y), size = 5, shape = 21, fill = "grey", color = "black") +
  geom_text(data = patches, aes(x = x, y = y, label = patch), vjust = -1) +
  geom_curve(data = movement_coords, aes(x = x_from, y = y_from,
                                         xend = x_to, yend = y_to),
             curvature = 0.2, arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
             color = "blue", size = 0.7) +
  labs(x = "X", y = "Y")
```


### Directed Dispersal to 'Patch 2'
```{r}
#DIAGRAM SHOWING DIRECT DISPERSAL TO PATCH 2
patches <- data.frame(
  patch = c("Patch 1", "Patch 2", "Patch 3"),
  x = c(1, 2, 3),
  y = c(1, 3, 2)
)

movement <- data.frame(
  from = c("Patch 1", "Patch 3", "Patch 1", "Patch 3"),
  to = c("Patch 2", "Patch 2", "Patch 2", "Patch 2")
)

movement_coords <- merge(movement, patches, by.x = "from", by.y = "patch")
movement_coords <- merge(movement_coords, patches, by.x = "to", by.y = "patch", suffixes = c("_from", "_to"))

ggplot() +
  geom_point(data = patches, aes(x = x, y = y), size = 5, shape = 21, fill = "grey", color = "black") +
  geom_text(data = patches, aes(x = x, y = y, label = patch), vjust = -1) +
  geom_curve(data = movement_coords[movement_coords$from == "Patch 1",], aes(x = x_from, y = y_from,
                                                                             xend = 2, yend = 3),
             curvature = 0.2, arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
             color = "blue", size = 0.7) +
  geom_curve(data = movement_coords[movement_coords$from == "Patch 1",], aes(x = x_from, y = y_from,
                                                                             xend = 2, yend = 3),
             curvature = -0.2, arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
             color = "blue", size = 0.7) +
  geom_curve(data = movement_coords[movement_coords$from == "Patch 3",], aes(x = x_from, y = y_from,
                                                                             xend = 2, yend = 3),
             curvature = 0.2, arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
             color = "blue", size = 0.7) +
  geom_curve(data = movement_coords[movement_coords$from == "Patch 3",], aes(x = x_from, y = y_from,
                                                                             xend = 2, yend = 3),
             curvature = -0.2, arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
             color = "blue", size = 0.7) +
  labs(x = "X", y = "Y")
```


### Population Dynamics
```{r}
# Define the model function
habitat_model <- function(time, N, params) {
  with(as.list(params), {
    dN <- c * N * (1 - N/T - (T-N)/T * (1/(v+1))) - e * N  
    return(list(dN))
  })
}

# Set parameters
c <- 0.125        # Colonization rate constant
e <- 0.05       # Rate at which patches become empty
T <- 100        # Carrying capacity
v <- 4        # Preference parameter

# Initial population size 
N0 <- 10        

# Time vector
times <- seq(0, 100, by = 0.1)

# Parameters list
params <- list(c = c, e = e, T = T, v = v)

# Solve the differential equation
output <- ode(y = N0, times = times, func = habitat_model, parms = params)

# Create a data frame for plotting
df <- as.data.frame(output)
colnames(df) <- c("time", "N")  # Rename columns if necessary

# Plot the population dynamics
ggplot(df, aes(x = time, y = N)) +
  geom_line() +
  labs(x = "Time", y = "Population size") +
  theme_minimal()
```


### Frequency Distribution of Population Dynamics
```{r}
par(mfrow= c(1,1))
habitat_model <- function(time, N, params) {
  with(as.list(params), {
    dN <- c * N * (1 - N/T - (T-N)/T * (1/(v+1))) - e * N  
    return(dN)
  })
}

# Set parameters
c <- 0.125      
e <- 0.05       
T <- 100       
v <- 4

# Set the range of N values
N_values <- seq(0, T, length.out = 1000)

# Calculate dN/dt for each N value
dN_dt_values <- habitat_model(NULL, N_values, list(c = c, T = T, e = e, v = v))
hist(dN_dt_values, breaks = 20, xlab = "dN/dt", ylab = "Frequency")
```


### Population Dynamics for Differing Colonisation Rates 'c'
```{r}
habitat_model <- function(time, N, params) {
  with(as.list(params), {
    dN <- c * N * (1 - N/T - (T-N)/T * (1/(v+1))) - e * N  
    return(list(dN))
  })
}
simulate_population <- function(colonization_rate) {
  # Set parameters
  params <- list(c = colonization_rate, e = 0.05, T = 100, v = 0.5)
  
  # Solve the differential equation
  output <- ode(y = N0, times = times, func = habitat_model, parms = params)
  
  # Data frame for plotting
  df <- as.data.frame(output)
  colnames(df) <- c("time", "N")
  df$colonization_rate <- colonization_rate
  
  return(df)
}

# Initial conditions and time vector
N0 <- 10
times <- seq(0, 100, by = 0.1)

# Simulate population dynamics for different colonization rates
colonization_rates <- c(0, 0.5, 1)
population_data <- lapply(colonization_rates, simulate_population)

population_df <- do.call(rbind, population_data)

# Plot population dynamics for different colonization rates
ggplot(population_df, aes(x = time, y = N, color = factor(colonization_rate))) +
  geom_line() +
  labs(x = "Time", y = "Population size", color = "Colonization rate") +
  theme_minimal()
```


### Population dynamics for differing preference rates 'v'
```{r}
# Population dynamics for a given preference parameter
simulate_population_v <- function(preference_parameter) {
  # Set parameters
  params <- list(c = 0.125, e = 0.05, T = 50, v = preference_parameter)
  
  # Solve the differential equation
  output <- ode(y = N0, times = times, func = habitat_model, parms = params)
  
  # Create a data frame for plotting
  df <- as.data.frame(output)
  colnames(df) <- c("time", "N")
  
  # Add preference parameter as a column for plotting
  df$preference_parameter <- preference_parameter
  
  return(df)
}

# Set preference parameter values
preference_parameters <- c(0, 1, 4)

# Simulate population dynamics for different preference parameter values
population_data_v <- lapply(preference_parameters, simulate_population_v)

# Combine population data for varying preference parameters into a single data frame
population_df_v <- do.call(rbind, population_data_v)

# Plot population dynamics for different preference parameter values
ggplot(population_df_v, aes(x = time, y = N, color = factor(preference_parameter))) +
  geom_line() +
  labs(x = "Time", y = "Population size", color = "Preference parameter") +
  theme_minimal()
```


### Population Dynamics of two Connected Patches
```{r}
habitat_model <- function(time, N, params) {
  with(as.list(params), {
    dN <- c * N * (1 - N/T - (T-N)/T * v/(v+1)) - e * N  
    return(list(dN))
  })
}

set.seed(123)  

tspan = 100
c = 2
T = 100
v = 0.95
extinction = 0.05

x1 = matrix(0,1, tspan) 
x2 = matrix(0,1, tspan) 
x1[1] = runif(1); 
x2[1] = runif(1);
for(t in 1:(tspan-1)){
  
  x1[t+1] = x1[t]*exp(c*(1-(x1[t]/T)))
  x2[t+1] = x2[t]*exp(c*(1-(x2[t]/T)))
  
  x1disp = v*x1[t+1] 
  x2disp = v*x2[t+1] 
  
  x1[t+1] = x1[t+1] + x2disp - x1disp
  x2[t+1] = x2[t+1] + x1disp - x2disp
  
}
par(mfrow=c(1,2))
plot(1:tspan, x1,type='l',col='red', xlab='Time', ylab='Abundance', main
     = 'Patch 1')

plot(1:tspan, x2,type='l',col='blue', xlab='Time', ylab='Abundance',
     main = 'Patch 2')
```



### Pearsons Correlation Coefficient 
#### Testing the strength of the relationship between the two patches
```{r}
correlation_test <- cor.test(x1, x2)
print(correlation_test)
```



